Erosion of somatic tissue identity with loss of the X-linked intellectual disability factor KDM5C

### 2024-08-13 submission/manuscript.Rmd
edit results 4

While many germline genes have roles in both the male and female germline, some display sex-biased expression or have functions unique to eggs and sperm. To comprehensively assess if germline gene sex corresponds with _Kdm5c_ mutant sex, we filtered our list of germline-enriched genes for egg (XX) and sperm (XY) biased genes. We defined sex-biased genes as those whose expression in the opposite sex, at any time point, is no greater than 20% of the gene's maximum expression in a given sex. This yielded `r nrow(subset(germgenes, sexBias == "XX"))` egg-biased, `r nrow(subset(germgenes, sexBias == "XY"))` sperm-biased, and `r nrow(subset(germgenes, sexBias == "unbiased"))` unbiased germline-enriched genes, which is consistent with the testis overall having a more unique transcriptome than the ovary[@liComprehensiveMouseTranscriptomic2017] (Figure 4H). We found egg, sperm, and unbiased germline genes were dyregulated in all _Kdm5c_ mutants (Figure 4I-L). _Kdm5c_ mutant male and female-specific germline DEGs were also not biased to the corresponding germ cell sex, indicating differences between male and female germline gene dysregulation not due to sex-specific activation of sperm or egg-specific transcriptional programs. These results demonstrate sex influences the degree of gene misexpression with loss of KDM5C, but not the sex of germ cell-enriched genes.

>>>>

While many germline genes act in both the male and female germline, some display sex-biased expression or have functions unique to eggs and sperm. To comprehensively assess whether germline gene sex corresponds with _Kdm5c_ mutant sex, we filtered our list of germline-enriched genes for egg (XX) and sperm (XY) biased genes. We defined sex-biased genes as those whose expression in the opposite sex, at any time point, is no greater than 20% of the gene's maximum expression in a given sex. This yielded `r nrow(subset(germgenes, sexBias == "XX"))` egg-biased, `r nrow(subset(germgenes, sexBias == "XY"))` sperm-biased, and `r nrow(subset(germgenes, sexBias == "unbiased"))` unbiased germline-enriched genes (Figure 4H), which is consistent with the testis overall having a more unique transcriptome than the ovary[@liComprehensiveMouseTranscriptomic2017]. We found egg, sperm, and unbiased germline genes were dyregulated in all _Kdm5c_ mutants (Figure 4I-L). Germline genes dysregulated exclusively in males and females were also not biased towards their corresponding germ cell sex. This indicates differences between male and female germline gene dysregulation is not due to sex-specific activation of sperm or egg-specific transcriptional programs. These results demonstrate that sex influences the degree of gene misexpression with KDM5C loss, but not the germ cell sex in which the germline genes originate.
